---
title: | 
  1 Data overview PacBio samples included <br> (Figure 1)

author: "Rotem Aharon"
date: today
params:
  data_path: "../../data/"
  project_path: "/Users/rotem/Code/GitHub/manuscript-and-analysis-code/"
  rscripts_path: "../rscripts/needed_functions.R"
  bambu_rds: "bambu_output/all_samples_counts/se_bambu_result.rds"
  output_path: "../../data/analysis_output/all_samples_counts/"
  figures_path: "figures/figure1/"
 
---

The data analysed here is publicly available from the SG-Nex database (Ying Chen et al. 2025). The SG-NEx data was accessed in June 2025 [registry.opendata.aws/sg-nex-data](https://registry.opendata.aws/sg-nex-data).
We started by downloading 107 .bam files (samples), which include read alignment to the human genome (hg38) for each of the replicates in each of the cell lines. 
'bambu' v3.8.3 with NDR=0.1 was used to identify and quantify the genes and transcripts in all of the samples together. This resulted in a SummarisedExperiment R object, which includes transcript counts, and full length read counts as separate assays. Gene counts were obtained by using bambu::transcriptToGeneExpression, which effectively adds the transcript counts for each gene. 
For each sample, library size was obtained by adding up all transcript counts. Samples with very low or very high library size  <400,000 or >40,000,000 were excluded, as well as those that had full length to all transcript ratio lower than 0.2. 
With the remaining 79 samples, MDS dimensional reduction plots were generated using edgeR::plotMDS for each of the assays independently, with top=2000, to account for the fact transcript counts are not independent of each other.

### Setup, libraries and parameters {.unnumbered .unlisted}
```{r setup, message = TRUE, warning = TRUE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE, warning = FALSE)
knitr::opts_chunk$set(tidy = 'formatR') 
```



```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse)
# analysis pkgs
library(limma)
library(edgeR)
library(SummarizedExperiment)
library(bambu)

# plotting tools
library(ggplot2)
#library(gridExtra) 
library(RColorBrewer)
#library(cowplot)
library(ggpubr)
library(kableExtra)
#library(patchwork)
source("~/Code/manuscript-and-analysis-code/src/rscripts/needed_functions.R")
```


```{r set-parameters}

data_path <- params$data_path
project_path <- params$project_path

rscripts_path <- params$rscripts_path
source(rscripts_path)

figures_path <- params$figures_path
output_path <- params$output_path
  
bambu_se_rds_file <- params$bambu_rds
bambu_rds_path <- paste(data_path,bambu_se_rds_file,sep="")

```

### Load bambu output {.unnumbered .unlisted}
SummarizedExperiment with bambu output is re-loaded here, then cell-line and preparation-protocols labels are obtained based on sample name. 


 
```{r load-bambu-output} 
data_path <- params$data_path

bambu_se <- readRDS( bambu_rds_path)


analysis_se <- bambu_se


colData(analysis_se)$cellType <- as.factor(strsplit2(colData(analysis_se)$name,split="_")[,2])

colData(analysis_se)$raw_protocol <- as.factor(strsplit2(colData(analysis_se)$name,split="_")[,3])


#define (collapsed) protocols:
colData(analysis_se)$protocol <- fct_collapse(colData(analysis_se)$raw_protocol,"PCRcDNA" = c("cDNA", "cDNAStranded"))


# define groups based on cellType and  (collapsed) protocols:
colData(analysis_se)$group <- as.factor(paste(colData(analysis_se)$cellType,colData(analysis_se)$protocol,sep = "_"))
analysis_se

```



# Library size (1a,1b)

For each sample, obtain library size and calculate the MDS dimensional reduction with edgeR for each assay:  Gene counts, Transcript counts, and full length reads only counts.

```{r calc-dim-reduction, message = FALSE, warning = FALSE}
analysis_se <- add_dimReductions(analysis_se,calc_mds = FALSE)


#MCF7-EV samples that were quantified by bambu in earlier stages are exluded from downstream analysis  
analysis_se_no_ev <- analysis_se[, colData(analysis_se)$cellType  !="MCF7-EV"]

```



## Library size based on Transcript counts 
```{r all-samples-libsize2}

libsize_low_lim=400000
  
libsize_high_lim=40000000
all_colData <- data.frame(colData(analysis_se_no_ev))

# plot all library sizes in a boxplot
f1a <- all_colData%>%ggplot(aes(cellType,log2(counts_libsize),color= protocol))+geom_boxplot() +
  geom_jitter(shape=16, position=position_dodge(width = 0.75))+
  geom_hline(yintercept = log2(libsize_low_lim)) +geom_hline(yintercept = log2(libsize_high_lim)) + 
  labs(x ="",y= "log2[library size (all transcript counts)") +
  theme(legend.position = "none")

f1a
```

Adjust plot limits for better representation of variability between samples
```{r f1a_ylim}

f1a <- f1a + coord_cartesian(ylim = c(14, 26.5))
f1a

```


## Full length to all count library size ratios

```{r f1b}


cell_line_colours <- RColorBrewer::brewer.pal(9, "Oranges")
# hex <- hue_pal()(6)
named_cell_line_colours <- list(
  "A549" = cell_line_colours[3],
  "H9" = cell_line_colours[4],
  "HEYA8" = cell_line_colours[5],
  "HepG2" = cell_line_colours[6],
  "Hct116" = cell_line_colours[7],
  "K562" = cell_line_colours[8],
  "MCF7" = cell_line_colours[9]
)



p2 <-all_colData%>%ggplot(aes(x=log2(counts_libsize),y=fullLengthCounts_libsize/counts_libsize,
                              color = cellType,shape = protocol) ) + 
  scale_shape_manual(values = c(16,17,4,15)) +
  geom_point(,size =3,stroke=1)+ 
  geom_vline(xintercept = log2(libsize_low_lim)) +
  annotate(geom="text",x=log2(libsize_low_lim-50000),y=.55, label="Library size lower limit",angle='90')+
  geom_hline(yintercept = 0.2) +
  annotate(geom="text",x=23,y=.22, label="Full-length / all transcripts = 0.2",angle='0') +
  theme(legend.position = "none")+ scale_color_manual(values= named_cell_line_colours) +
  labs(x="Library size (log2, all transcript counts)",  y="Library size ratio \n (Full length transcripts)/(all transcripts)")

 p2


```


Adjust plot limits for better representation of variability between samples

```{r f1b_xlim}


f1b <- p2+ coord_cartesian(xlim = c(16, 25))
f1b
```
# Filter and calculate MDS features for selected samples (1c-1h)
```{r select-samples}


selected_samples <- analysis_se_no_ev[,(colData(analysis_se_no_ev)$counts_libsize>libsize_low_lim & colData(analysis_se_no_ev)$counts_libsize<libsize_high_lim) & (colData(analysis_se_no_ev)$fullLengthCounts_libsize/colData(analysis_se_no_ev)$counts_libsize  > 0.2 ) ]


saveRDS(selected_samples,paste(output_path,"selected_samples_includePacbio.RDS",sep=""))
selected_samples <- add_dimReductions(selected_samples,top =2000)

selected_samples_all_colData <- data.frame(colData(selected_samples))
```
:::{.callout-note}
The total number of samples quantified by bambu is `r dim(all_colData)[1]`. 
After filtering out samples based on library size aand ratio of full length reads  (only include   `r libsize_low_lim` < transcripts library size < `r libsize_high_lim`,  and ratio of full length reads >0.2), there are `r dim(selected_samples_all_colData)[1]` samples included in the MDS plots presented in figure 1


:::

```{r selected_samples_MDS}




p3 <-selected_samples_all_colData%>%ggplot(aes(gene_counts_mds_1, gene_counts_mds_2))  %>% 
  mds_aes() + labs(x= "MDS_1",y= "MDS_2", title = "Gene counts") + scale_color_manual(values= named_cell_line_colours)

p4 <-selected_samples_all_colData%>%ggplot(aes(gene_counts_mds_3, gene_counts_mds_4))  %>%
  mds_aes() + labs(x= "MDS_3",y= "MDS_4", title = "Gene counts")

p5 <-  selected_samples_all_colData%>%ggplot(aes(counts_mds_1, counts_mds_2)) %>%
  mds_aes() + labs(x= "MDS_1",y= "MDS_2", title = "Transcript counts")

p6 <- selected_samples_all_colData%>%ggplot(aes(counts_mds_3, counts_mds_4))  %>%
  mds_aes() + labs(x= "MDS_3",y= "MDS_4", title = "Transcript counts")

p7 <-  selected_samples_all_colData%>%ggplot(aes(fullLengthCounts_mds_1, fullLengthCounts_mds_2)) %>%
  mds_aes() + labs(x= "MDS_1",y= "MDS_2", title = "Full length transcript counts")
                                
p8 <-  selected_samples_all_colData%>%ggplot(aes(fullLengthCounts_mds_3, fullLengthCounts_mds_4)) %>%
  mds_aes() + labs(x= "MDS_3",y= "MDS_4", title = "Full length transcript counts")
                                                                



 MDSplots <-ggarrange( 
  p3,p5,p7,p4 ,p6,p8, labels = c("c","e","g","d","f","h"),
           ncol = 3,nrow = 2,
  common.legend = TRUE, legend = "top"
  )
 
#MDSplots

```


##  Figure 1: Data overview

```{r save_figure}


  numPlots <- ggarrange( 
  f1a,   f1b, labels="auto",common.legend = TRUE,legend = "left",
  ncol = 2,nrow = 1
  )
 

pdf("allDataOverview_with_pacbio.pdf", width = 12, height = 12)
 dataOverview <- ggarrange(numPlots,MDSplots,
 nrow = 2,heights = c(1.2,3)          )
  dataOverview
  invisible(dev.off())
  

  png("allDataOverview_with_pacbio.png",width = 30, height = 30, units = "cm", 
    bg = "white",res = 300)
 dataOverview <- ggarrange(numPlots,MDSplots,
 nrow = 2,heights = c(1.2,3)          )
 
  dataOverview
  invisible(dev.off())

 
```
::: {layout-ncol=1}
![Figure 1:  a) Library size (number of mapped reads) in each sample, across preparation protocols and cell lines. The samples between the horizontal lines were retained for analysis. b) Fraction of full length transcript reads as a function of log2(library size). MDS plots of gene counts (c,d), total transcript counts (e,f) and full length transcript counts (g,h). Samples separate based on cell line (colour) but also on preparation protocol (shape) especially at the isoform level.   
](allDataOverview_with_pacbio.png)
:::
:::{.callout-note}
The total number of samples quantified by bambu is `r dim(all_colData)[1]`. 

After filtering out samples based on library size  (only include   `r libsize_low_lim` < transcripts library size < `r libsize_high_lim`), there are `r dim(selected_samples_all_colData)[1]` samples included in the MDS plots.
:::
# Appendix
## Samples list and exlusion criteria 
### Create table 

```{r}

downloaded_samples_table <- data.frame(colData(analysis_se))%>%dplyr::mutate(sample_name =name,
                                       library_size = counts_libsize,
                                       cell_line = cellType, grouped_protocol = protocol )%>% 
  dplyr::mutate (protocol = raw_protocol) %>% select(-c(raw_protocol,cellType,group,gene_counts_libsize))



downloaded_samples_table$in_1a_b <-  downloaded_samples_table$cell_line != "MCF7-EV"
downloaded_samples_table$in_1a_b <- ifelse(downloaded_samples_table$in_1a_b, "Yes", "No")



downloaded_samples_table$exluded_from_1c_h <-  downloaded_samples_table$cell_line == "MCF7-EV" |
  downloaded_samples_table$counts_libsize < libsize_low_lim | 
                  downloaded_samples_table$counts_libsize > libsize_high_lim | 
                  downloaded_samples_table$fullLengthCounts_libsize/downloaded_samples_table$counts_libsize < 0.2
downloaded_samples_table$in_1c_h <- ifelse(downloaded_samples_table$exluded_from_1c_h, "No", "Yes")


```
```{r}
ONTonly_samples <- "../../data/bambu_output/ont_only_counts/se_bambu_result.rds"

ont_all_quantification_results <- readRDS(ONTonly_samples)
ont_all_quantification_results <- add_dimReductions(ont_all_quantification_results,calc_mds = FALSE)
ont_all_quantified_samples <- colData(ont_all_quantification_results)

missinig_samples_names <- setdiff(ont_all_quantified_samples$name,downloaded_samples_table$name)
missinig_samples_names
```
```{r}



# Add values manually for now
downloaded_samples_table[missinig_samples_names,] <- c("SGNex_A549_cDNAStranded_replicate3_run3","cDNAStranded",NA,NA,"SGNex_A549_cDNAStranded_replicate3_run3",NA,"A549","PCRcDNA","No",TRUE,"No")


ont_all_quantified_samples <- ont_all_quantified_samples[,c("name","counts_libsize")]
colnames(ont_all_quantified_samples) <- c("name", "ONTrun_counts_libsize")


downloaded_samples_table <- downloaded_samples_table %>% full_join(data.frame(ont_all_quantified_samples),by =join_by("name"))


downloaded_samples_table$in_sup1a_b <- downloaded_samples_table$grouped_protocol != "PacBio-SMRTcell" &
                                      downloaded_samples_table$cell_line !="MCF7-EV"
  
downloaded_samples_table$in_sup1a_b <- ifelse(downloaded_samples_table$in_sup1a_b, "Yes", "No")

selected_samples_ont_only_se_path <- "../../data/analysis_output/ont_only_counts/selected_samples.rds"

ont_selected_samples_se<- readRDS( selected_samples_ont_only_se_path)
downloaded_samples_table$in_sup1c_h <- downloaded_samples_table$name %in% colData(ont_selected_samples_se)$name 
  



downloaded_samples_table$exluded_from_DTU <- ! downloaded_samples_table$in_sup1c_h |
                  all_colData$protocol == "PacBio-SMRTcell" | all_colData$protocol == "directcDNA" 

  
downloaded_samples_table$in_sup1c_h <- ifelse(downloaded_samples_table$in_sup1c_h,  "Yes","No")
downloaded_samples_table$in_DE_fig2 <- downloaded_samples_table$in_sup1c_h 
downloaded_samples_table$in_DTU_fig3 <- ifelse(downloaded_samples_table$exluded_from_DTU, "No", "Yes")

```
```{r}

downloaded_samples_table <- downloaded_samples_table%>% 
  select(-c(name,  exluded_from_1c_h,exluded_from_DTU, counts_libsize,fullLengthCounts_libsize ))  %>% 
  relocate(c(protocol,grouped_protocol), .before = cell_line) # %>% relocate(c(protocol,grouped_protocol), .before = cell_line)


downloaded_samples_table<- downloaded_samples_table[order(downloaded_samples_table$cell_line,downloaded_samples_table$protocol),]

# Count number of samples included in each column
yes_counts <- colSums(
downloaded_samples_table  == "Yes", na.rm = TRUE)

# Append to the bottom of the data frame
downloaded_samples_table <- rbind(
downloaded_samples_table , yes_counts)

#Give the new row a clear row name
rownames(downloaded_samples_table)[nrow(downloaded_samples_table)] <- "Total"


```



```{r sample_inclusion_table}


knitr::kable(
  x = downloaded_samples_table,
  caption = paste0("Sample inclusion table"), row.names =FALSE
)



```

## Auto format table to pdf (using latex)
```{r}
latex_table <- kable(downloaded_samples_table, row.names =FALSE,
                     format = "latex",
                     booktabs = TRUE,
                     longtable = TRUE,
                     escape = TRUE,
                     caption = "Sample inclusion table",
                     align = rep('c', ncol(downloaded_samples_table))) %>%  
  kable_styling(latex_options = c("striped","hold_position", "repeat_header"),
                font_size = 6,
                full_width = FALSE,
                position = "center")

# Use landscape geometry instead of kableExtra's landscape()
latex_doc <- paste0(
  "\\documentclass{article}\n",
  "\\usepackage[utf8]{inputenc}\n",
  "\\usepackage{booktabs}\n",
  "\\usepackage{longtable}\n",
  "\\usepackage{array}\n",
  "\\usepackage{colortbl}\n",
  "\\usepackage{xcolor}\n",
  "\\usepackage[landscape, margin=0.5in]{geometry}\n",
  "\\begin{document}\n",
  latex_table,
  "\n\\end{document}"
)

# Write LaTeX file
writeLines(latex_doc, "table1.tex")

# Compile to PDF (requires tinytex or LaTeX installation)
tinytex::pdflatex("table1.tex")

```


![Table 1: Sample inclusion table ](table1.pdf){width="100%"}

## Session info

```{r}
sessionInfo()
```